Data Analysis Extensions

### 2016-02-04 RFCs/HDF5_Library/AnalysisExt/analysis_ext-body.tex
More fixes to analysis RFC
Former-commit-id: c8ef4ae1fe2c15cc648c50526e6b8da0c5e65c7a

\section{Introduction}
When working on large datasets, finding and selecting the interesting pieces of
the data can be a cumbersome process. Currently, the HDF5 library enables the 
application developer to select, read and write data but does not provide any 
mechanism to select and retrieve pieces without prior knowledge of their
content, or without the developer to provide the exact data coordinates that
he is willing to access. To satisfy that need, one must be able
to issue queries by specifying a data selection criteria. These queries,
when applied to the data, can then generate a selection, which contains the coordinates
that satisfy the query---this may imply accessing the data and selecting it,
depending on the query condition satisfaction. To accelerate and facilitate this
process (i.e., so that the data no longer needs to be directly accessed), one can
use indexing techniques, which consist of generating an index and using that index
to answer the query selection criteria and find the matching elements.

We define in this RFC the components that can enable application developers to
create complex and high-performance queries on both metadata and data elements
within an HDF5 container and retrieve the results of applying those query
operations. Support for these operations can be defined via:
\begin{itemize}
\item New \textit{query} objects\footnote{Query objects are
\textit{in-memory} objects, which therefore do not modify the content of the
container.\label{fn:object}} and API routines, enabling the construction of
query requests for execution on HDF5 containers;
\item New \textit{index} objects and API routines, which allows the creation of
indices on the contents of HDF5 containers, to improve query performance.
\end{itemize}

@@

The core query API is composed of
two routines: \texttt{H5Qcreate} and \texttt{H5Qcombine}. \texttt{H5Qcreate}
creates new queries, by specifying an aspect of an HDF5 container, such as:
\begin{itemize}
\item \texttt{H5Q\_TYPE\_DATA\_ELEM} (data element);
\item \texttt{H5Q\_TYPE\_LINK\_NAME} (link name);
\item \texttt{H5Q\_TYPE\_ATTR\_NAME} (attribute name);
\item \texttt{H5Q\_TYPE\_ATTR\_VALUE} (attribute value);
\end{itemize}
as well as a match operator, such as:
\begin{itemize}
\item \texttt{H5Q\_MATCH\_EQUAL} ($=$);
\item \texttt{H5Q\_MATCH\_NOT\_EQUAL} ($\neq$);
\item \texttt{H5Q\_MATCH\_LESS\_THAN} ($\leq$);
\item \texttt{H5Q\_MATCH\_GREATER\_THAN} ($\geq$);
\end{itemize}
and a value for the match operator.
Created query objects can be serialized and deserialized using the \texttt{H5Qencode}
and \texttt{H5Qdecode} routines\footnote{Serialization/deserialization of queries
were introduced so that queries can be sent through the network.}, and their
content can be retrieved using the corresponding accessor routines.
\texttt{H5Qcombine} combines two query objects into a
new query object, using boolean operators such as:
\begin{itemize}
\item \texttt{H5Q\_COMBINE\_AND} ($\mathrm{\land}$);
\item \texttt{H5Q\_COMBINE\_OR} ($\mathrm{\lor}$).
\end{itemize}
Queries created with \texttt{H5Qcombine} can be used as input to further
calls to \texttt{H5Qcombine}, creating more complex queries.

For example, a single call to \texttt{H5Qcreate} could create a query object
that would match data elements in any dataset within the container that are
equal to the value ``$\mathrm{17}$''. Another call to \texttt{H5Qcreate} could create a
query object that would match link names equal to ``$\mathrm{Pressure}$''.
Calling \texttt{H5Qcombine} with the $\mathrm{\land}$ operator and those two query
objects would create a new query object that matched elements equal to ``$\mathrm{17}$''
in HDF5 datasets with link names equal to ``$\mathrm{Pressure}$''.
Creating the data analysis extensions to HDF5 using a \textit{programmatic
interface} for defining queries avoids defining a text-based query language
as a core component of the data analysis interface, and is more in keeping with
the design and level of abstraction of the HDF5 API.
The HDF5 data model is more complex than traditional database tables and a
simpler query model would likely not be able to express the kinds of queries
needed to extract the full set of components of an HDF5 container. A text-based
query language (or GUI) could certainly be built on top of the query API
defined here to provide a more user-friendly (as opposed to
\textit{developer-friendly}) query syntax like ``$\mathrm{Pressure = 17}$''.
However, we regard this as out-of-scope for now.

\subsection{Query Execution and Views}
Applying a query to an HDF5 container creates an HDF5 \textit{view}. A view is
an anonymous HDF5 group in a temporary memory container. A view group contains up to
three 1-D dataset objects with references into the contents of the HDF5 container
that the query was applied to:
one with \textit{object} references, one with \textit{region} references and
one with \textit{attribute} references.

@@

A new view is created by the
following \texttt{H5Qapply} routine, which applies a query to an HDF5 container,
group hierarchy, or individual object and returns the object ID of the newly created
group. The attributes, objects, and/or data regions referenced within a view's
datasets can be retrieved by further HDF5 dataset (H5D*) API calls.

@@

Although the created view is stored in memory, it can be persisted
by calling \texttt{H5Ocopy()} to copy the group (stored in a virtual
container) to a persistent container. For
coherency, a time stamp may also be attached to it so that its states has a meaning
compared to the state of the container that the query was applied to (as the
container may have been modified in the meantime). It may be useful in
that case to define different states to the view (\textit{dead} or
\textit{live}) so that the user knows whether the view is current or not.

@@

\begin{table}[ht]\footnotesize
\caption{Query combinations and associated result reference type.}
\label{tab:querycomb}
\begin{tabular}{ll} \toprule
Query &
Result Type\\ \midrule
\texttt{H5Q\_TYPE\_DATA\_ELEM} & Dataset~Region \\
\texttt{H5Q\_TYPE\_ATTR\_VALUE} & Attribute \\
\texttt{H5Q\_TYPE\_ATTR\_NAME} & Attribute \\
\texttt{H5Q\_TYPE\_LINK\_NAME} & Object \\
Dataset~Element $\mathrm{\land}$ Dataset~Element & Dataset~Region \\
Dataset~Element $\mathrm{\land}$ Attribute & Dataset~Region \\
Dataset~Element $\mathrm{\land}$ Object & Dataset~Region \\
Attribute $\mathrm{\land}$ Attribute & Attribute \\
Attribute $\mathrm{\land}$ Object & Object \\
Object $\mathrm{\land}$ Object & Object \\
Dataset~Element $\mathrm{\lor}$ Dataset~Element & Dataset~Region \\
Dataset~Element $\mathrm{\lor}$ Attribute & Combination \\
Dataset~Element $\mathrm{\lor}$ Object & Combination \\
Dataset~Element $\mathrm{\lor}$ Combination & Combination \\
Attribute $\mathrm{\lor}$ Attribute & Attribute \\
Attribute $\mathrm{\lor}$ Object & Combination \\
Attribute $\mathrm{\lor}$ Combination & Combination \\
Object $\mathrm{\lor}$ Object & Object \\
Object $\mathrm{\lor}$ Combination & Combination \\
Combination $\mathrm{\lor}$ Combination & Combination \\
Combination $\mathrm{\land}$ Dataset~Element & None \\
Combination $\mathrm{\land}$ Attribute & None \\
Combination $\mathrm{\land}$ Object & None \\
Combination $\mathrm{\land}$ Combination & None \\
\bottomrule
\end{tabular}
\end{table}

@@

{\lstsetc
\begin{lstlisting}
hid_t H5Qapply_multi(size_t loc_count, hid_t *loc_ids, hid_t query_id,
  unsigned *result, hid_t vcpl_id);
\end{lstlisting}
}

Location IDs specified may be entire containers, groups in a container's hierarchy
or individual datasets within a container. These locations may reside within the
same container, or in multiple containers. Additionally, the \texttt{H5Qapply\_str}
routine allows the user to provide container names instead of HDF5 locations:

{\lstsetc
\begin{lstlisting}
hid_t H5Qapply_str(size_t cont_count, const char *cont_name[], hid_t query_id,
  unsigned *result, hid_t vcpl_id);
\end{lstlisting}
}

As a direct consequence, one may want to pass queries that match a particular
container name. Therefore the \texttt{H5Q\_TYPE\_CONT\_NAME} query type can be
used to construct a query that matches this new condition. The matching
condition may be passed in the form of a file name or a regular
expression.

(Optional?) Apply query on all the containers of a given directory with \texttt{H5Qapply\_dir}
{\lstsetc
\begin{lstlisting}
hid_t H5Qapply_dir(const char *path, hid_t query_id, unsigned *result, hid_t vcpl_id);
\end{lstlisting}
}

Applying cross-container queries creates a view that contains references
to external locations (as opposed to the previous \texttt{H5Qapply} call where
references were internal). Therefore, the extended HDF5 reference object
also includes a container identifier so that, when the view result is
given back to the user, a subsequent call to \texttt{H5Rdereference} can return
a reference to an object within an externally referenced container\footnote{
As a result, the HDF5 reference API does not require any changes.}.

Parallel execution of the query optional? \textit{Pass hint for parallel execution
through property list.}

\section{Index Objects}
Index objects are designed to accelerate creation of views from query operations.
For example, if the ``$\mathrm{(link\_name = Pressure)}$
$\mathrm{\land}$ $\mathrm{(data\_element = 17)}$'' query
(previously described) is going to be either frequently executed or executed
on a container that contains significantly large datasets, indices
could be created in that container, which would speed up the
creation of views when querying for link names and for data element values.
Indices created for accelerating the  ``$\mathrm{(link\_name = Pressure)}$'' or
``$\mathrm{(data\_element}$ $\mathrm{= 17)}$'' queries
would also improve view creation for the more complex
``$\mathrm{(link\_name = Pressure)}$ $\mathrm{\land}$
$\mathrm{(data\_element = 17)}$'' query.
Although creating indices for metadata components of queries, such as link or
attribute names, is possible, we first focus on index creation for dataset elements,
as they represent the largest volume of data in typical HPC application usage of
HDF5. Queries with metadata components execute properly,
but are not able to be accelerated with an index currently.

The indexing API can work in conjunction with the view creation. When an \texttt{H5Qapply}
call is made on a given location, an index attached to any dataset queried
for element value ranges will be used to speed up the query process and return
a dataspace selection to the library for later use.
There are different techniques for creating data element indices, and the most
efficient method will vary depending on the type of the data that is to be
indexed, its layout, etc. A new interface for the HDF5 library that uses a
plugin mechanism is therefore defined.

@@

This interface is defined for adding third-party indexing plugins,
such as FastBit~\cite{Wu05}, ALACRITY~\cite{alacrity13}, etc.
The interface provides indexing plugins with efficient access to the contents of
the container for both the creation and the maintenance of indices. In addition,
the interface allows third-party plugins to create private data structures
within the container for storing the contents of the index.
The current API as well as the plugin interface are presented below:

@@

\begin{figure}
\input{pics/hdf5_index}
\caption{Index information (plugin ID and metadata) is stored along the object header.}
\label{fig:hdf5_index}
\vspace{-20pt}
\end{figure}

@@

Because the amount of space taken by the index cannot be directly retrieved by
the user (since the datasets storing the indices are known only by the plugin
itself), the \texttt{get\_size} callback can query the amount of space that
the index takes in the file and users may correspondingly query that information using the
\texttt{H5Xget\_size} routine.

@@

\begin{figure}
\input{plots/indexing_perf}
\caption{Indexing performance.}
\label{fig:indexing_perf}
\vspace{-20pt}
\end{figure}

@@

There are some existing limitations in the use of indices in the current
implementation: FastBit and ALACRITY do not support incremental updates,
an index is a shared resource for a dataset. Taken together, these conspire to
put limits on application updates to datasets with indices.
Additionally, because FastBit and ALACRITY do not allow incremental updates to
an index, each modification to an existing index forces the index to be entirely
rebuilt. The limitation in FastBit and ALACRITY will need to be addressed in
the base packages' implementation, so that incremental updates
to their index information can be made.

\subsubsection*{Questions}
Some questions are still open regarding the handling of indices:
\begin{enumerate}
%\item How could an index be built and queried in parallel for a dataset that already exists?
\item How to handle index updates when the specified index plugin is not available?
\textit{In traditional databases, stored procedures are saved with the data and
therefore available at any time, but that is not the case here.
We could mark the index as out of date and rebuild the index when the plugin is
available again.}
\item How to handle index queries when the specified index plugin is not
available? \textit{We could fallback to another plugin and do a brute force query on the data.}
\end{enumerate}

@@

In a simple scenario, the HDF5 datatype used for creating the dataset can be defined
as a native and simple type. Therefore, building an index on this dataset implies
building that index from the entire dataset. However, in more complex scenarios, the
dataset may have been created by using a compound datatype, hence defining
multiple fields composed of native and simple datatypes within that same dataset.
Consequently, creating an index from that dataset requires the user to select a
particular field to be indexed, which may lead to having multiple indices per dataset
depending on the number of fields that it contains. This can be done by passing
the \texttt{datatype\_id} of the field to be indexed to the \texttt{xcpl\_id},
the index creation property list, of the \texttt{H5Xcreate} call, which passes
it down to the plugin \texttt{create} callback. As multiple fields can be defined,
the field \texttt{datatype\_id} must be stored along with the existing metadata, within the
\texttt{idx\_info} message (see~\FigureRef{fig:hdf5_index}) so that the index
associated to the field can be retrieved at the time of the query.
When doing a query, the compound type is passed to the \texttt{H5Qcreate} call.
The corresponding index is then used and the query is passed to the \texttt{query}
callback of the plugin.

@@

Existing HDF5 tools must be compatible and take into account the existence of
indices in the file if there are any. For reference, the following behavior for
the tools is given:

>>>>

\section{Introduction}
When working on large datasets, finding and selecting the interesting pieces of
the data can be a cumbersome process. Currently, the HDF5 library enables the 
application developer to select, read and write data but does not provide any 
mechanism to select and retrieve pieces without prior knowledge of its
content, or without the developer to provide the exact data coordinates that
he is willing to access. To satisfy that need, one must be able
to issue queries by specifying a data selection criteria. These queries,
applied to the data, can then generate a selection, which in turn contains a
set of coordinates
satisfying the query. However, generating a selection may actually imply accessing the
data. To accelerate and facilitate that
process (i.e., so that the actual data no longer needs to be accessed), one can
also generate indexes and use these indexes
to directly answer the specified query, by finding the coordinates of the matching
elements directly from the index.

We define in this RFC the components that can enable application developers to
create complex and high-performance queries on both metadata and data elements
within an HDF5 container and retrieve the results of applying those query
operations. Support for these operations can be defined via:
\begin{itemize}
\item New \textit{query} objects\footnote{Query objects are
\textit{in-memory} objects, which therefore do not modify the content of the
container.\label{fn:object}} and API routines, enabling the construction of
query requests for execution on HDF5 containers;
\item New \textit{index} objects and API routines, which allows the creation of
indexes on the contents of HDF5 containers, to improve query performance.
\end{itemize}

@@

The core query API is composed of
two routines: \texttt{H5Qcreate} and \texttt{H5Qcombine}. \texttt{H5Qcreate}
creates new queries, by specifying an aspect of an HDF5 container, such as:
\begin{itemize}
\item \texttt{H5Q\_TYPE\_DATA\_ELEM} (data element);
\item \texttt{H5Q\_TYPE\_LINK\_NAME} (link name);
\item \texttt{H5Q\_TYPE\_ATTR\_NAME} (attribute name);
\item \texttt{H5Q\_TYPE\_ATTR\_VALUE} (attribute value);
\end{itemize}
as well as a match operator, such as:
\begin{itemize}
\item \texttt{H5Q\_MATCH\_EQUAL} ($=$);
\item \texttt{H5Q\_MATCH\_NOT\_EQUAL} ($\neq$);
\item \texttt{H5Q\_MATCH\_LESS\_THAN} ($\leq$);
\item \texttt{H5Q\_MATCH\_GREATER\_THAN} ($\geq$);
\end{itemize}
and a value for the match operator.
Created query objects can be serialized and deserialized using the \texttt{H5Qencode}
and \texttt{H5Qdecode} routines\footnote{Serialization/deserialization of queries
were introduced so that queries can be sent through the network.}, and their
content can be retrieved using the corresponding accessor routines.
\texttt{H5Qcombine} combines two query objects into a
new query object, using boolean operators such as:
\begin{itemize}
\item \texttt{H5Q\_COMBINE\_AND} ($\mathrm{\land}$);
\item \texttt{H5Q\_COMBINE\_OR} ($\mathrm{\lor}$).
\end{itemize}
Queries created with \texttt{H5Qcombine} can be used as input to further
calls to \texttt{H5Qcombine}, creating more complex queries.
For example, a single call to \texttt{H5Qcreate} could create a query object
that would match data elements in any dataset within the container that are
equal to the value ``$\mathrm{17}$''. Another call to \texttt{H5Qcreate} could create a
query object that would match link names equal to ``$\mathrm{Pressure}$''.
Calling \texttt{H5Qcombine} with the $\mathrm{\land}$ operator and those two query
objects would create a new query object that matched elements equal to ``$\mathrm{17}$''
in HDF5 datasets with link names equal to ``$\mathrm{Pressure}$''.
Creating the data analysis extensions to HDF5 using a \textit{programmatic
interface} for defining queries avoids defining a text-based query language
as a core component of the data analysis interface, and is more in keeping with
the design and level of abstraction of the HDF5 API.
The HDF5 data model is more complex than traditional database tables and a
simpler query model would likely not be able to express the kinds of queries
needed to extract the full set of components of an HDF5 container. A text-based
query language (or GUI) could certainly be built on top of the query API
defined here to provide a more user-friendly (as opposed to
\textit{developer-friendly}) query syntax like ``$\mathrm{Pressure = 17}$''.
However, we regard this as out-of-scope for now.

\subsection{Query Execution and Views}
Applying a query to an HDF5 container creates an HDF5 \textit{view}. A view is
an anonymous HDF5 group created in a temporary memory container. A view group contains up to
three 1-D dataset objects with references into the contents of the HDF5 container
that the query was applied to:
one with \textit{object} references, one with \textit{region} references and
one with \textit{attribute} references.

@@

A new view is created by the
following \texttt{H5Qapply} routine, which applies a query to an HDF5 container,
group hierarchy, or individual object and returns the object ID of the newly created
group. The attributes, objects, and/or data regions referenced within a view's
datasets can be retrieved by further HDF5 dataset (\texttt{H5D*}) API calls.

@@

Although the created view is stored in memory, it can be persisted
by calling \texttt{H5Ocopy()} to copy the group (stored in a virtual
container) to a persistent container. For
coherency, a time stamp may also be attached to it so that its states has a meaning
compared to the state of the container that the query was applied to (as the
container may have been modified in the meantime). It may be useful in
the future to define different states to the view (\textit{dead} or
\textit{live}) so that the user knows whether the view is current or not.

@@

\begin{longtable}{ll}
\vspace{-30pt}\\
\caption{Query combinations and associated result reference type.}\\
\label{tab:querycomb} \vspace{-10pt} \\
\toprule
Query & Result Type\\ \midrule
\endfirsthead
\toprule
Query & Result Type\\ \midrule
\endhead
\multicolumn{2}{c}{\ldots}\\
\bottomrule
\endfoot
\bottomrule
\endlastfoot
\texttt{H5Q\_TYPE\_DATA\_ELEM} & Dataset~Region \\
\texttt{H5Q\_TYPE\_ATTR\_VALUE} & Attribute \\
\texttt{H5Q\_TYPE\_ATTR\_NAME} & Attribute \\
\texttt{H5Q\_TYPE\_LINK\_NAME} & Object \\
Dataset~Element $\mathrm{\land}$ Dataset~Element & Dataset~Region \\
Dataset~Element $\mathrm{\land}$ Attribute & Dataset~Region \\
Dataset~Element $\mathrm{\land}$ Object & Dataset~Region \\
Attribute $\mathrm{\land}$ Attribute & Attribute \\
Attribute $\mathrm{\land}$ Object & Object \\
Object $\mathrm{\land}$ Object & Object \\
Dataset~Element $\mathrm{\lor}$ Dataset~Element & Dataset~Region \\
Dataset~Element $\mathrm{\lor}$ Attribute & Combination \\
Dataset~Element $\mathrm{\lor}$ Object & Combination \\
Dataset~Element $\mathrm{\lor}$ Combination & Combination \\
Attribute $\mathrm{\lor}$ Attribute & Attribute \\
Attribute $\mathrm{\lor}$ Object & Combination \\
Attribute $\mathrm{\lor}$ Combination & Combination \\
Object $\mathrm{\lor}$ Object & Object \\
Object $\mathrm{\lor}$ Combination & Combination \\
Combination $\mathrm{\lor}$ Combination & Combination \\
Combination $\mathrm{\land}$ Dataset~Element & None \\
Combination $\mathrm{\land}$ Attribute & None \\
Combination $\mathrm{\land}$ Object & None \\
Combination $\mathrm{\land}$ Combination & None \\
\end{longtable}

@@

{\lstsetc
\begin{lstlisting}
hid_t H5Qapply_multi(size_t loc_count, hid_t loc_ids[], hid_t query_id,
  unsigned *result, hid_t vcpl_id);
\end{lstlisting}
}

Location IDs specified may be entire containers, groups in a container's hierarchy
or individual datasets within a container. These locations may reside within the
same container, or in multiple containers.
Applying cross-container queries creates a view that contains references
to external locations (as opposed to the previous \texttt{H5Qapply} call where
references were internal). Therefore, the extended HDF5 reference object
also includes a container identifier so that, when the view result is
given back to the user, a subsequent call to \texttt{H5Rdereference} can return
a reference to an object within an externally referenced container\footnote{
Please refer to the H5R reference RFC.}.

%Additionally, the \texttt{H5Qapply\_str}
%routine allows the user to provide container names instead of HDF5 locations:
%{\lstsetc
%\begin{lstlisting}
%hid_t H5Qapply_str(size_t cont_count, const char *cont_name[], hid_t query_id,
%  unsigned *result, hid_t vcpl_id);
%\end{lstlisting}
%}
%As a direct consequence, one may want to pass queries that match a particular
%container name. Therefore the \texttt{H5Q\_TYPE\_CONT\_NAME} query type can be
%used to construct a query that matches this new condition. The matching
%condition may be passed in the form of a file name or a regular
%expression.
%(Optional?) Apply query on all the containers of a given directory with \texttt{H5Qapply\_dir}
%{\lstsetc
%\begin{lstlisting}
%hid_t H5Qapply_dir(const char *path, hid_t query_id, unsigned *result, hid_t vcpl_id);
%\end{lstlisting}
%}

%Parallel execution of the query optional? \textit{Pass hint for parallel execution
%through property list.}

\section{Index Objects}
Index objects are designed to accelerate creation of views from query operations.
For example, if the previously described ``$\mathrm{(link\_name = Pressure)}$
$\mathrm{\land}$ $\mathrm{(data\_element = 17)}$'' query
is going to be either frequently executed or executed
on a container that contains significantly large datasets, indexes
could be created in that container, which would speed up the
creation of views when querying for link names and for data element values.
Indexes created for accelerating the  ``$\mathrm{(link\_name = Pressure)}$'' or
``$\mathrm{(data\_element}$ $\mathrm{= 17)}$'' queries
would also improve view creation for the more complex
``$\mathrm{(link\_name = Pressure)}$ $\mathrm{\land}$
$\mathrm{(data\_element = 17)}$'' query.
Although creating indexes for metadata components of queries, such as link or
attribute names, is possible, we first focus on index creation for dataset elements,
as they represent the largest volume of data in typical HPC application usage of
HDF5. Queries with metadata components execute properly,
but are not able to be accelerated with an index currently.

The indexing API can work in conjunction with the view creation. When an \texttt{H5Qapply}
call is made on a given location, an index attached to any dataset queried
for element value ranges will be used to speed up the query process and return
a dataspace selection to the library for later use.
There are different techniques for creating data element indexes, and the most
efficient method will vary depending on the type of the data that is to be
indexed, its layout, etc. A new interface for the HDF5 library that uses a
plugin mechanism is therefore defined.

@@

This interface is defined for adding third-party indexing plugins,
such as FastBit~\cite{Wu05}, ALACRITY~\cite{alacrity13}, etc.
The interface provides indexing plugins with efficient access to the contents of
the container for both the creation and the maintenance of indexes. In addition,
the interface allows third-party plugins to create private data structures
within the container for storing the contents of the index.
The current API as well as the plugin interface are presented below:

@@

\begin{figure}
\vspace{-30pt}
\input{pics/hdf5_index}
\caption{Index information (plugin ID and metadata) is stored along the object header.}
\label{fig:hdf5_index}
\vspace{-20pt}
\end{figure}

@@

Because the amount of space taken by the index cannot be directly retrieved by
the user (since the datasets storing the indexes are known only by the plugin
itself), the \texttt{get\_size} callback can query the amount of space that
the index takes in the file and users may correspondingly query that information using the
\texttt{H5Xget\_size} routine.

@@

\begin{figure}
\vspace{-10pt}
\input{plots/indexing_perf}
\caption{Indexing performance.}
\label{fig:indexing_perf}
\vspace{-20pt}
\end{figure}

@@

There are some existing limitations in the use of indexes in the current
implementation: FastBit and ALACRITY do not support incremental updates,
an index is a shared resource for a dataset. Taken together, these conspire to
put limits on application updates to datasets with indexes.
Additionally, because FastBit and ALACRITY do not allow incremental updates to
an index, each modification to an existing index forces the index to be entirely
rebuilt. The limitation in FastBit and ALACRITY will need to be addressed in
the base packages' implementation, so that incremental updates
to their index information can be made.

\subsubsection*{Questions}
Some questions are still open regarding the handling of indexes:
\begin{enumerate}
%\item How could an index be built and queried in parallel for a dataset that already exists?
\item How to handle index updates when the specified index plugin is not available?
\textit{In traditional databases, stored procedures are saved with the data and
therefore available at any time, but that is not the case here.
We could mark the index as out of date and rebuild the index when the plugin is
available again.}
\item How to handle index queries when the specified index plugin is not
available? \textit{We could fallback to another plugin and do a brute force query on the data.}
\end{enumerate}

@@

In a simple scenario, the HDF5 datatype used for creating the dataset can be defined
as a native and simple type. Therefore, building an index on this dataset implies
building that index from the entire dataset. However, in more complex scenarios, the
dataset may have been created by using a compound datatype, hence defining
multiple fields composed of native and simple datatypes within that same dataset.
Consequently, creating an index from that dataset requires the user to select a
particular field to be indexed, which may lead to having multiple indexes per dataset
depending on the number of fields that it contains. This can be done by passing
the \texttt{datatype\_id} of the field to be indexed to the \texttt{xcpl\_id},
the index creation property list, of the \texttt{H5Xcreate} call, which passes
it down to the plugin \texttt{create} callback. As multiple fields can be defined,
the field \texttt{datatype\_id} must be stored along with the existing metadata, within the
\texttt{idx\_info} message (see~\FigureRef{fig:hdf5_index}) so that the index
associated to the field can be retrieved at the time of the query.
When doing a query, the compound type is passed to the \texttt{H5Qcreate} call.
The corresponding index is then used and the query is passed to the \texttt{query}
callback of the plugin.

@@

Existing HDF5 tools must be compatible and take into account the existence of
indexes in the file if there are any. For reference, the following behavior for
the tools is given:

### 2016-02-04 RFCs/HDF5_Library/AnalysisExt/analysis_ext.tex
More fixes to analysis RFC
Former-commit-id: c8ef4ae1fe2c15cc648c50526e6b8da0c5e65c7a

@@

% Title, author, etc
\title{Data Analysis Extensions}
\author{Jerome Soumagne}
\author{Quincey Koziol}
\date{Oct 15, 2015}
\rfcversion{2014-07-17.v7}
\revision{Jul. 17, 2014}{Version 1 circulated for comment within The HDF Group.}
\revision{Aug. 20, 2014}{Version 2 circulated for comment within The HDF Group.}
\revision{Nov. 14, 2014}{Version 3 circulated for comment within The HDF Group.}
\revision{Feb. 25, 2015}{Version 4 edits and typo fixing.}
\revision{Jul. 24, 2015}{Version 5 edits to view object.}
\revision{Oct. 14, 2015}{Version 6 rework of query/view sections.}
\revision{Oct. 15, 2015}{Version 7 first draft of cross-container queries.}

@@

%% Abstract
\begin{abstract}
Accessing, selecting and retrieving data from an HDF5 container can be a time
consuming process, particularly so when data is very large. To enable, ease and 
accelerate this process, we introduce in this RFC extensions to the library
to efficiently query, select and index data.
\end{abstract}

>>>>

% Break table
\usepackage{longtable}

@@

% Title, author, etc
\title{Data Analysis Extensions}
\author{Jerome Soumagne}
\author{Quincey Koziol}
\date{Feb 3, 2016}
\rfcversion{2014-07-17.v9}
\revision{Jul. 17, 2014}{Version 1 circulated for comment within The HDF Group.}
\revision{Aug. 20, 2014}{Version 2 circulated for comment within The HDF Group.}
\revision{Nov. 14, 2014}{Version 3 circulated for comment within The HDF Group.}
\revision{Feb. 25, 2015}{Version 4 edits and typo fixing.}
\revision{Jul. 24, 2015}{Version 5 edits to view object.}
\revision{Oct. 14, 2015}{Version 6 rework of query/view sections.}
\revision{Oct. 15, 2015}{Version 7 first draft of cross-container queries.}
\revision{Oct. 31, 2015}{Version 8 some corrections.}
\revision{Feb. 3, 2016}{Version 9 more text corrections.}

@@

%% Abstract
\begin{abstract}
Accessing, selecting and retrieving data from an HDF5 container can be a time
consuming process, particularly so when data is very large. To enable, ease and 
accelerate this process, we introduce a set of extensions to the HDF5 library
that enable efficient query, selection, and indexing of data. This document defines
the \texttt{H5Q} \textit{query} and \texttt{H5X} \textit{indexing} interfaces as
well as a \textit{view} structure that can be used to store and retrieve query results.
\end{abstract}
